---
title: "Cleaning and Wrangling Data"
author: "Ted Flynn"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, include=FALSE}
# Load necessary functions ----
library(dplyr)
library(tidyr)
library(readr)
```

# Introduction
About the data: This data was downloaded [here](https://knb.ecoinformatics.org/view/df35b.304.2) on April 4, 2023. 

## Read data
```{r read data, warning=FALSE, message=FALSE}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore data
```{r explore}
colnames(catch_original)

head(catch_original)

summary(catch_original)

sort(unique(catch_original$Year))

unique(catch_original$Region)
```

# Clean and Reshape Data

## Selecting and Removing Columns with `select()`


```{r select}

catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Coho, Pink, Chum)

head(catch_data)

# This does the same thing as the code above
catch_data <- catch_original %>% 
  select(-All, -notesRegCode)

head(catch_data)

```

```{r checking}
glimpse(catch_data) # Shows that Chinook is categorized as a character var

```

## Changing column content using `mutate()`
```{r changing data type}
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)
```

## Checking NAs

```{r checking na}

i <- which(is.na(catch_clean$Chinook))

catch_data[i,] # One row has an I instead of 1

```

## Fixing Chinook Column 
```{r fix chinook}
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I", 
                           "1",
                           Chinook),
         Chinook = as.numeric(Chinook))
```

## Reshape data

```{r reshape}

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")

head(catch_long)

```

From long to wide use `pivot_wider()`.

```{r reshape wide}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)

head(catch_wide)

```

## Renaming Columns
```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)

colnames(catch_long)
```

## Add a column using `mutate()`

```{r adding column}

catch_long <- catch_long %>% 
  mutate(catch = catch_thousands*1000) %>% 
  select(-catch_thousands)

head(catch_long)

```

## Split-Apply-Combine Workflow
We are going to use the `group_by()` and `summarize()` functions.

```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_region)

mean_reg_yr <- catch_long %>% 
  group_by(Region, Year) %>% 
  summarize(catch_mean = mean(catch))

head(mean_reg_yr)

n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(n = n())

head(n_region)
```

## Filtering by rows with `filter()`

```{r}

sse_catch <- catch_long %>% 
  filter(Region == "SSE")

head(sse_catch)
```

## Clean Data all at Once

```{r}
catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
                               method = "libcurl"))

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))

head(mean_region)
```


## Remove columns


## Check column types
## Reshape data
