---
title: "Intro to Data Visualization"
author: "Ted Flynn"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r load packages, warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)


```

## Read in data
```{r read data, warning=FALSE, message=FALSE}
escapement <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")
```

## Checking data
```{r checking}
## Check out column names
colnames(escapement)

## Peak at each column and class
glimpse(escapement)

## From when to when
range(escapement$sampleDate)

## How frequent?
head(escapement$sampleDate)
tail(escapement$sampleDate)

## Which species?
unique(escapement$Species)

```


## Getting data ready to plot

```{r data prep}
annual_esc <- escapement %>%
    separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>%
    mutate(Year = as.numeric(Year)) %>%
    group_by(Species, SASAP.Region, Year) %>%
    summarize(escapement = sum(DailyCount)) %>%
    filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

unique(annual_esc$Species)
head(annual_esc)
  
```

# Plotting data with `ggplot`

## Essential components
```{r essential}
## Options 1
ggplot(data = annual_esc,
       aes(x = Species,
           y = escapement)) +
  geom_col()

## Option 2
ggplot(data = annual_esc)+
       geom_col(aes(x = Species,
                    y = escapement))

## Option 3
ggplot() +
  geom_col(data = annual_esc,
           aes(x = Species,
               y = escapement))
```

## Different geoms

**Boxplot**
```{r boxplot}
annual_esc %>% 
  filter(Year == 1974,
         Species %in% c("Chum","Pink")) %>% 
  ggplot(aes(x = Species,
             y = escapement)) +
  geom_boxplot(width = 0.4)


```

**Line and Point**
```{r line plot}
annual_esc %>% 
  filter(Species == "Sockeye",
         SASAP.Region == "Bristol Bay") %>% 
  ggplot(aes(x = Year,
             y = escapement)) +
  geom_line() +
  geom_point()
```

## Customizing Plots
fill = "blue" outside the `aes()`
```{r customization}
ggplot(annual_esc,
       aes(x = Species,
           y = escapement,
           fill = "blue"))+
  geom_col()

```

Adding color by region
```{r color by region}
ggplot(annual_esc,
       aes(x = Species,
           y = escapement,
           fill = SASAP.Region))+
  geom_col()
```

## Plotting multiple graphs
```{r subset by year}
## Subset with data from years 2000 to 2016

annual_esc_2000s <- annual_esc %>%
    filter(Year %in% c(2000:2016))

## Quick check
unique(annual_esc_2000s$Year)
```

## Plot with Regions
```{r region multiplot}
ggplot(annual_esc_2000s,
       aes(x = Year,
           y = escapement,
           color = Species)) +
    geom_line() +
    geom_point() +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y")
```

```{r plotting with themes}
ggplot(annual_esc_2000s,
       aes(x = Year,
           y = escapement,
           color = Species)) +
    geom_line() +
    geom_point() +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    theme(legend.position = "bottom",
          legend.title = element_blank())
```

### Saving a theme as an object
```{r saving theme}

my_theme <- theme()+
  theme(legend.position = "bottom",
        legend.title = element_blank())

```

Create plot with saved theme. 
```{r plot saved theme}
ggplot(annual_esc_2000s,
       aes(x = Year,
           y = Species,
           color = Species))+
  geom_line() +
  geom_point() +
  facet_wrap(~SASAP.Region,
             scales = "free_y") +
  ylab("Escapement") +
  ggtitle("Annual Salmon Escapement by Region") +
  my_theme
```


Setting scales using the `scales()` package
```{r scales}
ggplot(annual_esc_2000s,
       aes(x = Year,
           y = escapement,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    scale_y_continuous(labels = comma) +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))
```

Angle years to 45 degrees and x axis breaks by 2
```{r plot angle text}
annual_region_plot <- ggplot(annual_esc_2000s,
                             aes(x = Year,
                                 y = escapement,
                                 color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    scale_y_continuous(labels = comma) +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    xlab("\nYear") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))

annual_region_plot
```

### Saving Plots
```{r}
ggsave("figs/annual_esc_region.png", 
       annual_region_plot, 
       width = 12,
       height = 8,
       units = "in")
```

## Maps with `leaflet()`

```{r making maps}
location <- escapement %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()
```

Create map with `leaflet()`

```{r create maps}
leaflet(location) %>% 
  addTiles() %>% 
  addMarkers()
```

```{r}
leaflet(location) %>%
    addWMSTiles(
        "https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0",
        layers = 'GEBCO_LATEST',
        attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net"
    ) %>%
    addCircleMarkers(
        lng = ~ Longitude,
        lat = ~ Latitude,
        popup = ~ Location,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1
    )
```



